Researchers collaborate with same-gendered colleagues more often than expected across the life sciences

### 2019-03-04 manuscript/homophily_manuscript.tex
Ready for plos resubmission

\usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath}
\usepackage{booktabs} \usepackage{caption}
\usepackage[labelfont=bf]{caption} \usepackage{titling}
\pretitle{\begin{flushleft}\LARGE}
\posttitle{\par\end{flushleft}\vskip 0.5em}
\preauthor{\begin{flushleft}\large \lineskip 0.5em}
\postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
\postdate{\par\end{flushleft}} \usepackage{fancyhdr} \pagestyle{fancy}
\fancyhead[LO,LE]{\textsl{\leftmark}} \rhead[]{Gender and coauthorship}
\usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot}
\usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}
\linenumbers

@@

Writing papers, networking, and collaboration are all instrumental to
research productivity and academic career advancement {[}22--25{]}, and
dozens of studies have tested for gender differences in these areas
{[}5,26--29{]}. For example, studies have concluded that women tend to
be less involved in international collaboration {[}19,28,30--32{]},
collaborate less within their own university departments {[}31{]}, have
less prestigious collaborations {[}33{]}, and fewer collaborations in
total {[}34{]}. These gender differences in collaboration practice
presumably have multiple causes, which might include implicit and
explicit gender bias {[}20{]}, differential family obligations
{[}33,35,36{]}, gender differences in confidence or self-esteem
{[}37{]}, concerns relating to sexual harassment {[}38{]}, and unequal
access to conferences {[}39{]} or travel funds {[}32{]}.

A high, steadily increasing proportion of research papers is written by
more than one author {[}3{]}, making collaboration a key predictor of
publication output, and thus of career prospects {[}40,41{]}.
Additionally, empirical studies imply that mixed-gender or otherwise
diverse teams produce better outputs on collaborative tasks than less
diverse teams {[}42--48{]}. For reasons such as these, multiple studies
have examined the author lists of published research articles in order
to test for gender differences in collaboration frequency or pattern. To
our knowledge, most or all such studies imply that men co-publish with
men, and women with women, more often than expected if collaborators
assort randomly with respect to gender {[}49--58{]}. This pattern of
assortative publishing has often been termed `gender homophily'.

However, we believe that earlier studies of gender homophily were
hindered by a largely unacknowledged statistical issue that we will
refer to as the Wahlund effect (\autoref{wahlund_plot}), by analogy with
the conceptually similar Wahlund effect in population genetics {[}59{]}.
The Wahlund effect makes it deceptively difficult to infer gender-based
preferences simply by counting the number of same- and mixed-gender
coauthorships. Essentially, whenever coauthorship data are sampled from
two or more discrete sets of literature, which vary in the author gender
ratio and which are largely not connected by collaboration, the number
of same-gendered coauthors will be inflated. This can give the
impression that authors preferentially publish with same-gendered
colleagues even if no gender preferences exist, or if the true
preference is for opposite-gendered colleagues (`gender heterophily').
For example, a sample of literature containing a mixture of
bioinformatics and cell biology papers will probably contain an excess
of mostly-male and mostly-female author lists, simply because
researchers usually collaborate within their own discipline, and because
the author gender ratio is more male-biased in bioinformatics than in
cell biology {[}5{]}.

@@

There was no indication that journals publishing on a wide range of
topics have higher \(\alpha'\) values than more specialised journals,
due to the Wahlund effect. For example, the journal category
`Multidisciplinary' -- which includes journals like \emph{PLOS ONE},
\emph{Nature}, \emph{Science}, and \emph{PNAS} -- did not have markedly
elevated \(\alpha'\) (\autoref{alpha_histograms}). This result suggests
that our estimates of homophily, and estimates from some of the earlier
studies listed in the Introduction, are probably not inflated by the
presence of disparate research topics (with variable author gender
ratios) being published within individual journals.

@@

An important limitation of our study is that we cannot reliably
determine the cause(s) of the observed excess of same-gender
coauthorships. As well as the obvious interpretation -- conscious or
unconscious selection of same-gender collaborators by men, by women, or
both -- our results could be partly explained by uncontrolled Wahlund
effects. However, we suspect the contribution of these uncontrolled
artefacts to be minor, for four reasons: we found positive \(\alpha'\)
after controlling for three obvious sources of Wahlund effect; there was
no inflation of \(\alpha'\) in highly multidisciplinary journals
relative to specialised journals; restricting the data by country
yielded similar estimates of \(\alpha'\); and we used a simple model to
show that differences in gender ratio between career stages are unlikely
to fully explain our results. On balance, we believe the data suggest
that it is likely that some researchers preferentially select
same-gendered collaborators, although it is difficult to ascertain what
proportion of people show such a preference, or how much the strength of
the preference varies between individuals. We also note that even in a
world in which collaboration was completely random with respect to
gender, a high proportion of individual researchers would have entirely
same-gendered collaborators by chance alone (especially in gender-biased
disciplines); thus, individuals who only have same-gendered co-authors
are not necessarily doing anything differently from people with
gender-balanced co-authors.

>>>>

\usepackage{microtype} \usepackage[all]{hypcap} \usepackage{amsmath} \usepackage{booktabs} \usepackage{caption} \usepackage[labelfont=bf]{caption} \usepackage{titling}

\pretitle{\begin{flushleft}\LARGE}
\posttitle{\par\end{flushleft}\vskip 0.5em}
\preauthor{\begin{flushleft}\large \lineskip 0.5em}
\postauthor{\par\end{flushleft}} \predate{\begin{flushleft}\large}
\postdate{\par\end{flushleft}}

\usepackage{fancyhdr}

\pagestyle{fancy} \fancyhead[LO,LE]{\textsl{\leftmark}}
\rhead[]{Gender and coauthorship}

\usepackage[noabbrev,capitalise]{cleveref} \usepackage{titlefoot} \usepackage{amssymb} \usepackage{rotating} \usepackage{lineno}

\linenumbers

@@

Writing papers, networking, and collaboration are all instrumental to
research productivity and academic career advancement {[}22--25{]}, and
dozens of studies have tested for gender differences in these areas
{[}5,26--29{]}. For example, studies have concluded that women tend to
be less involved in international collaboration {[}19,28,30--32{]},
collaborate less within their own university departments {[}31{]}, have
less prestigious collaborations {[}33{]}, and fewer collaborations in
total {[}34{]}. These gender differences in collaboration presumably
have multiple causes, which might include implicit and explicit gender
bias {[}20{]}, differential family obligations {[}33,35,36{]}, gender
differences in confidence or self-esteem {[}37{]}, concerns relating to
sexual harassment {[}38{]}, and unequal access to conferences {[}39{]}
or travel funds {[}32{]}.

A high, steadily increasing proportion of research papers is written by
more than one author {[}3{]}, making collaboration a key predictor of
publication output, and thus of career prospects {[}40,41{]}.
Additionally, empirical studies imply that mixed-gender or otherwise
diverse teams produce better outputs on collaborative tasks than less
diverse teams {[}42--48{]}. For reasons such as these, multiple studies
have examined the author lists of published research articles in order
to test for gender differences in collaboration frequency or pattern. To
our knowledge, most or all such studies imply that men co-publish with
men, and women with women, more often than expected if collaborators
assort randomly with respect to gender {[}49--58{]}. This non-random
assortment is often termed `gender homophily'.

We believe that most or all earlier studies of gender homophily were
hindered by a largely unacknowledged statistical issue that we will
refer to as the Wahlund effect (\autoref{wahlund_plot}), by analogy with
the conceptually similar Wahlund effect in population genetics {[}59{]}.
The Wahlund effect makes it deceptively difficult to infer gender-based
preferences simply by counting the number of same- and mixed-gender
coauthorships. Essentially, whenever coauthorship data are sampled from
two or more discrete sets of literature, which vary in the author gender
ratio and which are largely not connected by collaboration, the number
of same-gendered coauthors will be inflated. This can give the
impression that authors preferentially publish with same-gendered
colleagues even if no gender preferences exist, or if the true
preference is for opposite-gendered colleagues (`gender heterophily').
For example, a sample of literature containing a mixture of
bioinformatics and cell biology papers will probably contain an excess
of mostly-male and mostly-female author lists, simply because
researchers usually collaborate within their own discipline, and because
the author gender ratio is more male-biased in bioinformatics than in
cell biology {[}5{]}.

@@

There was no indication that journals publishing on a wide range of
topics have higher \(\alpha'\) values than more specialised journals,
due to the Wahlund effect. For example, the journal category
`Multidisciplinary' -- which includes journals like \emph{PLOS ONE},
\emph{Nature}, \emph{Science}, and \emph{PNAS} -- did not have markedly
elevated \(\alpha'\) (\autoref{alpha_histograms}). This result suggests
that our estimates of homophily, and estimates from some of the earlier
studies listed in the Introduction, are probably not markedly inflated
by the presence of disparate research topics (with variable author
gender ratios) being published within individual journals.

@@

An important limitation of our study is that we cannot reliably
determine the cause(s) of the observed excess of same-gender
coauthorships. As well as the obvious interpretation -- conscious or
unconscious selection of same-gender collaborators by men, by women, or
both -- our results could be partly explained by uncontrolled Wahlund
effects. However, we suspect the contribution of these uncontrolled
artefacts to be minor, for four reasons: we found positive \(\alpha'\)
after controlling for three obvious sources of Wahlund effect; there was
no inflation of \(\alpha'\) in highly multidisciplinary journals
relative to specialised journals; restricting the data by country
yielded similar estimates of \(\alpha'\); and we used a simple model to
show that differences in gender ratio between career stages are unlikely
to fully explain our results. On balance, we believe the data suggest
that it is likely that some researchers preferentially select
same-gendered collaborators, although it is difficult to ascertain what
proportion of people show such a preference, or how much the strength of
the preference varies between individuals. We also note that even in a
world in which everyone selected their collaborators at random with
respect to gender, a high proportion of individual researchers would
have entirely same-gendered collaborators by chance alone (especially in
gender-biased disciplines); thus, individuals who only have
same-gendered co-authors are not necessarily doing anything differently
from people with gender-balanced co-authors.
